\documentclass[a4paper, 12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[brazil]{babel}
\renewcommand{\footnotesize}{\fontsize{8pt}{10pt}\selectfont} %mmm: diminui o tamanho da fonte de footnote
\usepackage{titling} % Customizing the title section
\setlength{\droptitle}{-6\baselineskip} % Move the title up

\usepackage{geometry}
\geometry{left=2cm, right=2cm, bottom=2.5cm, top=3cm} %diminuir a margem

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{setspace}
% COLOR %Para escrever observações de cores diferentes
\usepackage{color}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\seunome}[1]{\textcolor[rgb]{0.29,0.77,0.83}{#1}}
\newcommand{\nome}[1]{\textcolor[RGB]{138,43,226}{#1}}

\singlespacing

\newcommand{\pyline}[1]{\mintinline{python}|#1|}
\newcommand{\yell}[1]{{\color{blue} [#1]}}
\newcommand{\tom}[1]{\yell{#1 --tom}}

\title{Integrating Proof Production in Cylindrical Algebraic Decomposition}
\author{Anonymous}
\date{}

\begin{document}

\maketitle

\vspace{-20pt}
\section{Introduction}
\subsection{Context}

% Motivation

\qquad One of the most fundamental properties of any software program is its correctness with respect
to a given specification. This ensures that the program behaves as expected for all possible inputs.
For example, one possible specification for a sorting routine is that its output list is always sorted.
Thus, it will be considered correct if this property is respected for any input list it receives.
Similarly, specifications can be defined for far more complex systems, such as airplane
controllers or medical devices, in which correctness is necessary for guaranteeing its safety.
Therefore, the problem of \textit{verifying} if a piece of software is correct is critical in the modern world.

Indeed, there exists a number of examples of malfunctioning software causing irreparable damage
that fortify the importance of verification: the explosion of the rocket Ariane 5 in 1996, caused by
a incorrect conversion between numerical values; the faulty operation of Knight Capital's trading algorithm,
leading to a debt of 440 milion dollars for the company in 2012; the deaths of at least 3 people caused by
the Therac-25 radiation therapy machine, which was attributed to the existence of race conditions in its
software; the out-of-bounds read performed by CloudStrike's software, resulting in milions of Windows machines
becoming temporarily unusable in 2024.

% Formal methods

There are a variety of techniques to obtain correctness evidence for a
software. The most common one is the development of tests, and
there are many types of bugs that can be
discovered with its execution. In fact, this approach is enough for a large
amount of problems that are solved by software engineering.
However, in general, tests can't guarantee that a program doesn't have flaws, since the number of
valid inputs is almost always exceedingly large or infinite.
In this context, a promising alternative is to carry out the formal verification
of the system, that is, the rigorous checking that it meets the predetermined specification.
This is achieved by reasoning symbolically about the code, which enables validating the specification
for all possible inputs at once.

Automatic theorem provers (ATPs) are systems in which the user enters as input a conjecture written in
a logical
formalism and the system responds with ``yes'' or ``no'', depending on whether that conjecture is true or
not (or ``unknown'' if it could not get to a conclusion).
One remarkable use case of ATPs is to formally verify software systems by encoding
their behavior and the desired property as logical formulas. More concretely,
one can write a formula that expresses the idea ``there exists some configuration
that forces my system into a given bad state'' and send it to the ATP. If it responds
with ``yes'', it will usually also present the undesired configuration. Otherwise,
if it responds with ``no'', the user can be sure that the bad state will never occur,
assuming he trusts the ATP.

\subsubsection{SMT Solvers}

There are many classes of automatic theorem provers. Even though the techniques we aim
to develop in this project could potentially serve any of them, we will be mainly concerned with
their integration with Satisfiability Modulo Theories (SMT) solvers~\cite{smtbook}.
Satisfiability Modulo Thoeries is an extension of the Boolean Satisfiability Problem (SAT),
in which the underlying logic system is First-Order Logic. Additionally, a set of theories
is included, that is, a set of domains over which the input terms can be defined (as opposed
to just Booleans), as well as operators from these domains. For instance, the following is
a valid SMT formula: $x \ge 0 \wedge x < 1 \wedge f(x) \neq f(0)$, where $x$ is an integer
and $f$ is a function from integers to integers.
%
SMT solvers are systems specialized in determining automatically whether such formulas are satisfiable,
in the sense that there exist a valuation for the variables such that, after substituting the variables by
their values, the formula reduces to \textit{true}.

The division of the logic into theories makes it possible to use specialized procedures for each
one of them. This contributed vastly to the success of SMT solvers both in formal verification of software and in other domains.

\subsection{Generating Proofs}
One important concern over ATPs is whether or not they are, themselves, correct. This is crucial
in the context of formal verification since any mistake in their logic or implementation will invalidate any
correctness result they produce. Unfortunately, since these solvers are usually highly complex,
it is remarkably hard to prove that they are completely correct. Moreover, even if a proof is concluded,
any modification in the system will require its correctness to be proven again. One alternative that has been
gaining attention in recent years~\cite{generatingProofs, proofsInSmt} is to modify the system to produce
proof certificates on the fly of its results. This means that, in addition to outputting ``yes'' or ``no'', they will
also provide a witness that serves as a precise logical argument supporting their conclusion. These witnesses can be
checked independently by a third-party software, transferring the need of trusting the solver (which often
comprises hundreds of thousands of lines of code) into a need to trust the proof checker (which usually is
written in a few thousands of lines of code).

A second application of this modification is to provide a powerful mean for debugging the solver.
In many cases~\cite{generatingProofs}, if there is a bug in the ATP, it will produce an incorrect
proof (or fail to produce one) and this proof will give hints to the developers about where the
bug probably is. In addition to that, the process of doing this modification forces
the developers to understand deeply the implementation of the solver.

Finally, a third application of instrumenting solvers to produce proofs is exporting them to
other contexts where they are highly valuable, such as in proof assistants.
%
Proof assistants are interactive systems in which the user states a theorem and manually constructs
a proof for it. The system verifies each step of the proof, providing feedback to the user, and only
accepts the proof if all steps are correct.
%
Similar to proof checkers, proof assistants are easier to be trusted because the part of their code
responsible for checking proofs is relatively small.
%
One remarkable use case of interactive provers is the formalization of mathematics.
Since the formalization of the Four Color Theorem~\cite{fctCoq} in a proof assistant called Coq,
mathematicians have used more and more these tools to increase the level of certainty
they have in their results, and also, to facilitate large scale collaboration between them.
More recently, in 2022, the fields-medalist Peter Scholze announced that he and his collaborators
completed the formalization of Perfectoid Spaces~\cite{scholze}, a result that lies in the
cutting edge of research in mathematics, using another proof assistant.
%
However, these tools are more difficult to use, as the user must construct their proofs
manually. Additionally, these proofs need to be much more detailed than traditional pen-and-paper proofs,
since they must be automatically verified by a computer.
%
One notable way of improving the usability of proof assistants is to connect them with an ATP,
making it possible for the user to discharge some proof obligations to the automatic solver.
This approach has been proven to be successful in the past~\cite{coq2, reconstruct}, but it
only works with ATPs that are capable of producing proofs.
%
All these applications go to show the importance of proofs for automatic theorem provers.


\subsection{Solving Arithmetic Constraints}

The problem of solving a set of arithmetic constraints (i.e.\ expressions that set conditions
on the values of variables according to arithmetic relationships) is,
not surprisingly, pervasive when trying to prove conjectures automatically. In the general case, these expressions
can also contain quantifiers binding the variables. Solving this problem requires
performing \textit{quantifier elimination}~\cite{mayeroQE}, that is, finding a new formula without quantifiers
that is equivalent to the original one. Since this kind of problem is ubiquitous in theorem
proving, it is crucial for automatic theorem provers to implement efficient methods for solving it.


The inherent difficulty of this problem varies depending on the characteristics of the formula:
if the problem is linear (i.e.\ does not contain multiplication of free variables), then
it can be solved in polynomial time if all variables range over the real numbers~\cite{Kha79}, and it
is NP-complete otherwise~\cite{Kar72}; when non-linear restrictions are introduced, the complexity increases
dramatically, as it requires an algorithm with doubly-exponential ($\mathcal{O}(2^{2^{n}})$) time complexity
to solve it if all variables range over reals~\cite{computerAlgebraBook},
and it is undecidable otherwise~\cite{integerUndec}.

While solving such constraints is already complex, an additional challenge emerges in generating
formal proofs that verify the correctness of the solutions.
%
For linear constraints with real variables there are established techniques
that both solve the constraints and produce proofs of correctness with high reliability.
%
The standard strategy in this case is to use a decision procedure based on the simplex method~\cite{simplex_dpllt}, which can either produce a valuation for the variables if the formula
is satisfiable, or the Farkas' coefficients~\cite{farkas_ref} otherwise.
%
These coefficients have the property of producing a constraint that is impossible to satisfy
($1 < 0$, for instance) when they are used to scale the original constraints and the resulting
expressions are added together.
%
If the original formula is unsatisfiable, the Farkas' lemma guarantee that such coefficients always
exist, thus, they can be safely used to build a proof showing that there is no solution for the problem.
%
Constraints with integer variables pose an additional challenge, since the algorithm for solving them
requires the usage of the Branch and Bound technique combined with the simplex method. The proofs produced
in this case have to cover the reasoning behind this method, but it is also well understood how one
can modify the algorithm to generate them~\cite{challengesArithProofs}.

The situation becomes substantially more difficult when we consider constraints that permit
multiplication of variables.
For the case with only real variables, the most efficient approach known is to use an algorithm based on the
Cylindrical Algebraic Decomposition (CAD)~\cite{col75},
which requires an enormous amount of computational power to be constructed. Even though
this approach has been successful in handling non-linear constraints, there is currently
no comprehensive framework for generating verifiable proofs of its results.
%
As we will show in Section~\ref{sec:relatedWork}, there have been some works in this direction, but they
target algorithms that are either far less efficient than CAD or that deals with a simplification of this
problem.
%
This gap presents a significant limitation, as non-linear reasoning is essential both for verification of
software and for formalization of mathematics.

\subsection{Goals}

The goal of this future thesis is to bridge this gap by modifying the CAD algorithm to provide
proofs of correctness, advancing the reach and reliability of automatic theorem proving
for non-linear arithmetic. Once the algorithm is designed, the next goal is to integrate it
into the Lean~\cite{lean4} proof assistant. We chose Lean due to its growing acceptance in the mathematical
community, and also for its rich mathematical library, which already covers a large portion that
we expect to be necessary for the execution of this work. Lastly, we will integrate our algorithm
in the cvc5~\cite{cvc5} SMT solver. This tool is a good fit for our project since its architecture
was already adapted to support proof production~\cite{flexibleProofs}. It is important to remark
that the machinery we plan to implement for these integrations can potentially be reused in other proof
assistants and SMT solvers by adapting it to their contexts.

In Section~\ref{sec:methodology} we give more details about the challenges of our goals and how we
plan to tackle them.


\section{Related Work}
\label{sec:relatedWork}

There are mainly two ways to certify the results of an algorithm: one of them is to modify the
algorithms to produce proofs, as we have been discussing; the other one is to write a formal proof in a proof assistant
showing why the algorithm is correct. Historically, in the context of quantifier elimination for real arithmetic, the latter have
been more explored than the former. There have been a number of proofs of correctness for a variety of algorithms related to CAD.
However, one important remark is that all of them are either incomplete, solve a more restricted version of the problem
or are significantly less efficient than CAD.

The methods described in~\cite{kosaian2,nipkowVS,chaiebVS} present proofs for the \textit{Virtual Substitution} method.
This is an efficient incomplete method that works for constraints where the corresponding polynomials have low degree.
In particular, the implementation of~\cite{nipkowVS} deals with polynomials of degree at most $1$, the one
in~\cite{chaiebVS} deals with degree at most $2$ as long as all constraints are equalities and~\cite{kosaian2}
deals with any constraints with degree at most $2$.

One particularly well-explored variation of the problem is restricting expressions to only contain a single quantified variable.
There exists a simpler version of the CAD algorithm that only works in this case. The work in~\cite{univCad} formally verifies
this version of the algorithm in the Isabelle/HOL proof assistant, making use of the Mathematica system to discharge
part of the computational work, which highly optimizes their method. This cooperation is done in an untrusted way, meaning
that the computation done by Mathematica have to be rechecked in proof assistant, thus not compromising the safety of the tool at
the cost of decreasing its efficiency.

Complete algorithms for the unrestricted version of the problem tend to be highly complex. In particular, it has been
proven~\cite{doublyExponential} that any algorithm that solves it will require at least doubly exponential
($\mathcal{O}(2^{2^{n}})$) time, where $n$ is the number of quantified variables. The work
in~\cite{coqTarski} uses the Coq proof assistant to formalize Tarski's algorithm, which was the first one to completely
solve real quantifier elimination. Similarly, in~\cite{mcLaughlin} is presented a proof producing version of another
complete procedure, known as Cohen-H{\"o}rmander algorithm. While these works bring an extraordinary theoretical
value, they can only be used in very small cases, as both of the algorithms have non-elementary complexity, meaning that
their worst-case running time cannot be bounded by any tower of powers of two. This presents a significant disadvantage
in comparison to the cylindrical algebraic decomposition, which has a doubly-exponential complexity. It is important to
remark that the CAD algorithm is considerably more complicated than both Tarski's and Cohen-H{\"o}rmander algorithm.
Additionally, it relies deeply on a more involved mathematical theory, which potentially increases the difficulty
on formalizing it or modifying it for proof-production.

\section{Methodology}
\label{sec:methodology}

In order to modify an algorithm to produce proofs, it is necessary to understand the logic behind
each of its details. Thus, the first step of this project will be to study in depth the
theory behind the Cylindrical Algebraic Decomposition. This will be achieved by reading of books in the topic,
such as~\cite{computerAlgebraBook}, reading documentations of open-source systems that use this
algorithm, such as SageMath~\cite{sage} and discussing with researchers that have worked with it in the past.
The latter will be facilitated by the fact that the execution of this project will be supported by the
development team of cvc5.

Next, the existing implementations of the algorithm will be compared, and the most suitable one for the
first prototype will be identified. In principle, we beleive that the Cylindrical Algebraic
Coverings~\cite{coverings}, which only deals with formulas where all variables are existentially quantified,
would be a good fit, since it is considerably simpler than CAD while still being suitable for SMT solving. Additionally, there has been
some work~\cite{proofsCac} suggesting that this could be indeed a simpler task and
outlining some possible directions for executing it.

The next milestone will be to implement a prototype. For that, we have to define a set of proof steps
that comprehend the logical reasoning behind the algorithm. On one hand, it is important that
these proof steps aren't too coarse-grained, in the sense that it should not be hard for a program to check a particular application
of one of these proof steps;
on the other hand, they still need to be expressive enough to explain why the algorithm made a specific decision.
We will use as a baseline the description of the proof rules that cvc5 uses for its algorithms~\footnote{Which can be found at: \href{https://cvc5.github.io/docs/cvc5-1.0.0/proofs/proof\_rules.html}{https://cvc5.github.io/docs/cvc5-1.0.0/proofs/proof\_rules.html}}.
After that, we will implement a new version of the algorithm that effectively constructs a proof with these rules during its execution.

Once the algorithm is completed, we will integrate it in the Lean proof assistant by implementing, within its framework,
each one of the defined proof rules. This implementation permits us to parse and reconstruct proofs generated by the algorithm
in Lean, in a way that enables the proof assistant to recognize the unsatisfatibility of the
original non-linear problem as a theorem. This step will also validate whether the set of proof
rules we defined is appropriate or not: if they are too coarse-grained, it will be very hard to implement them
in Lean, thus we would have to redesign them.

Finally, we will integrate the algorithm in the SMT solver cvc5. Since cvc5 already has a (non-proof producing) solver for
non-linear arithmetic, it should not be hard to replace their module by our algorithm. Once this is done, we will adapt the
algorithm to fit into cvc5's proof production architecture.


\section{Schedule}

The following table shows the plan of courses to be taken in each
semester:

\begin{table}[!htbp]
\caption[Schedule]{Initial Schedule of Classes}
\vspace{5pt}
\centering
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}p{2cm}p{14cm}@{}}
\toprule
\textbf{Semester} & \textbf{Classes} \\
\midrule
2025/01
   & Otimiza\c{c}\~ao Combinat\'ortia \\
   & Teoria e Pr\'atica de Resolu\c{c}\~ao SMT \\
\hline
2025/02
   & Est\'agio em Doc\^encia \\
   & \'Algebra Avan\c{c}ada (Departamento de Matem\'atica) \\
\hline
  2026/01
  & Programa\c{c}\~ao N\~ao Linear \\
  & T\'opicos em Otimiza\c{c}\~ao \\
\hline
  2026/02
  & Est\'agio em Doc\^encia \\
  & Programa\c{c}\~ao Paralela \\

 \bottomrule
\end{tabular}
\end{table}

The next table presents the planning of activities that will be
performed per semester related to the
project development:

\renewcommand{\refname}{References}
\bibliographystyle{plain}
\bibliography{references}
\end{document}
